Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-03-01 analysis/paper/paper.Rmd
minor edit to 1st paragraph intro; added a few notes

@@

Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015]. Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat. Although the local extinction of species from individual fragments is common *(CITATIONS_1)*, the demographic mechanisms underlying these extinctions are poorly understood *(CITATIONS_2)*. However, it is often assumed that that the dramatically altered abiotic conditions often observed in fragments can depress rates of reproduction or survivorship *(CITATIONS_3)*, leading to local or regional population declines *(CITATIONS_2)*. This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates. For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in extensive fragmentation and creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008].

@@

Populations in fragmented tropical landscapes are also at risk from climate change *(CITATIONS_5)*. These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhi2008] and increases in both drought frequency and severity [@caiIncreasingFrequencyExtreme2014] *(add CITATIONS_6)*. All of these have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable - especially for plants, which are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999]). Alternatively, populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@betts2019].

@@

@@

## Focal species

>>>>

<!-- Consider co-authorship of Cris Jurinitz, Maria Rosa Darrigo, Paulo Rubim, John Kress. -->

@@

Deforestation driven by the expansion of agriculture and other human activities is a primary driver of biodiversity loss worldwide [@haddadHabitatFragmentationIts2015]. Deforestation also results in landscapes where the remaining forest can be highly fragmented, with remnants of different sizes embedded in a matrix of often highly-contrasting habitat. Although the local extinction of species from individual fragments is common *(CITATIONS_1)*, the demographic mechanisms underlying these extinctions are poorly understood *(CITATIONS_2)*. However, it is often assumed that that the dramatically altered environmental conditions often observed in fragments can depress rates of reproduction or survivorship *(CITATIONS_3)*, leading to local or regional population declines *(CITATIONS_2)*. This is thought to be particularly true in species-rich tropical forests [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999], which continue to be cleared and fragmented at unprecedented rates. For example, over 24 million ha of the Brazilian Amazon have been cleared in the last two decades *(CITATION_4)*, resulting in extensive fragmentation and creation of over 70,000 km of new forest edges annually [@broadbentForestFragmentationEdge2008]. Despite this, evidence linking fragmentation-related changes in environmental conditions with population-level demographic responses remains scarce.

@@

Populations in fragmented tropical landscapes are also at risk from climate change *(CITATIONS_5)*. These effects may be especially pronounced in Amazonia, parts of which have been experiencing a drying trend since the mid-1970s and where climate models for the 21st century predict substantial decreases in dry season precipitation [@malhi2008] and increases in both drought frequency and severity [@caiIncreasingFrequencyExtreme2014] *(add CITATIONS_6)*. All of these have the potential to exacerbate the microclimatic changes observed in forest fragments that are hypothesized to make populations demographically vulnerable in fragments - especially for plants, which unlike animals are unable to move in response to increased in air temperature, decreased relative humidity, reduced soil moisture, or the increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999]). Alternatively, populations persisting in fragments may be more demographically resilient to drought than those in primary forest if the individuals susceptible to desiccation died quickly after fragment isolation [@betts2019].

@@

<!-- Fig 1. (map) can go in the Supplement Materials -->

@@

<!-- Rainfall data should prob be a figure in main text, since the paper deals with rainfall and seasonality. I’d think of citing the (eventual) BDFFP rain package -->
## Focal species
